Capstone Movielens Report

### 2025-05-04 reports/capstone-movielens-report.draft4.Rmd
Draft report updated.

---
title: "Capstone Movielens Report"
author: "Azamat Kurbanaev"
date: "`r Sys.Date()`"
bibliography: references.bib
#csl: bit-numerical-mathematics.csl 
envir:
  parent.frame()
output:
  pdf_document: 
    df_print: paged
    toc: true
    fig_caption: true
    keep_tex: true
    citation_package: biblatex
colorlinks: true
linkcolor: red
citecolor: blue
header-includes:
    - \usepackage{hyperref}
    - \usepackage[natbib=true, 
                  style=numeric, 
                  backref=true, 
                  sorting=none]{biblatex}
    - \hypersetup{backref,
          pdfpagemode=Normal,
          colorlinks=true,
          implicit=false}
vignette: >
  %\VignetteIndexEntry{RMarkdown Citations - Numeric Style}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::knitr}
---

@@

## Plot dependency of RMSEs vs Overal Mean Rating Deviation
\
```{r}

data.frame(delta = deviation, 
           delta.RMSE = deviation.RMSE) |> 
tuning.plot(title = TeX(r'[RMSE as a function of deviation ($\delta$) from the Overall Mean Rating ($\hat{mu}$)]'),
              xname = "delta", 
              yname = "delta.RMSE", 
              xlabel = TeX(r'[$\delta$]'), 
              ylabel = "RMSE")

@@

### Datasets

``` {r , echo = FALSE}
```

@@

Let's install the development version of this package from the GitHub repository and attach the correspondent library to the global environment:
```{r }
if(!require(edx.capstone.movielens.data)) pak::pak("AzKurban-edX-DS/edx.capstone.movielens.data")

library(edx.capstone.movielens.data)
edx <- edx.capstone.movielens.data::edx
final_holdout_test <- edx.capstone.movielens.data::final_holdout_test

summary(edx)
summary(final_holdout_test)
```
#### `edx` Dataset
\
Let's look into the details of the `edx` dataset:
``` {r echo = TRUE}
str(edx)
```
Note that we have 9000055 rows and six columns in there:  
```{r }
dim_edx <- dim(edx)
print(dim_edx)
```

Also, we can see that no movies have a rating of 0. Movies are rated from 0.5 to 5.0 in 0.5 increments:
```{r }
#library(dplyr)
s <- edx |> group_by(rating) |>
  summarise(n = n())
print(s)
```

@@

## Methods / Analysis
:::{.callout-note}
All the source code of the R-scripts is available on the project's [GitHub repository](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/tree/main/r/src). 
:::

@@

Let's define some helper functions for logging and time-measuring features that we will use in our R scripts. Some of them are listed below:

```{r}
open_logfile <- function(file_name){
  log_file_name <- as.character(Sys.time()) |> 
    str_replace_all(':', '_') |> 
    str_replace(' ', 'T') |>
    str_c(file_name)

@@

:::{.callout-note}
The full source code of these functions is available in the [Logging Helper function section of the Capstone MovieLens Main R Script](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/capstone-movielens.main.R#L1).
:::

### Preparing train and set datasets
\
First, let's note that we have 10677 different movies: 
```{r}
n_movies <- n_distinct(edx$movieId)
print(n_movies)
```
and 69878 different users in the dataset:
```{r}
n_users <- n_distinct(edx$userId)
print(n_users)
```

Taking into consideration these observations, we came up with a decision to use the data from the dataset only for users who have provided at least 100 ratings.

Now let's split the `edx` dataset into a training set, which we will use to build and train our models, and a test set in which we will compute the accuracy of our predictions, the way described in the [Section 23.1.1 Movielens data](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_ mentioned above[@IDS2_23-1-1]:
```{r}
# Let's ignore the data for users who have not provided at least 100 ratings:
edx100 <- edx |> 
  group_by(userId) |>
  filter(n() >= 100) |>
  ungroup()

>>>>

---
title: "Capstone Movielens Report"
author: "Azamat Kurbanaev"
date: "`r Sys.Date()`"
bibliography: references.bib
#csl: bit-numerical-mathematics.csl 
envir:
  parent.frame()
output:
  pdf_document: 
    df_print: paged
    toc: true
    fig_caption: true
    keep_tex: true
    citation_package: biblatex
colorlinks: true
linkcolor: red
citecolor: blue
header-includes:
    - \usepackage{awesomebox}
    - \usepackage{hyperref}
    - \usepackage[natbib=true, 
                  style=numeric, 
                  backref=true, 
                  sorting=none]{biblatex}
    - \hypersetup{backref,
          pdfpagemode=Normal,
          colorlinks=true,
          implicit=false}
vignette: >
  %\VignetteIndexEntry{RMarkdown Citations - Numeric Style}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::knitr}
---

\newenvironment{infobox}[1]
  {
  \begin{itemize}
  \renewcommand{\labelitemi}{
    \raisebox{-.7\height}[0pt][0pt]{
      {\setkeys{Gin}{width=3em,keepaspectratio}
        \includegraphics{images/#1}}
    }
  }
  \setlength{\fboxsep}{1em}
  \begin{blackbox}
  \item
  }
  {
  \end{blackbox}
  \end{itemize}
  }

@@

@@

### Datasets Overview

@@

Let's install the development version of this package from the GitHub repository and attach the correspondent library to the global environment:
```{r eval=FALSE }
if(!require(edx.capstone.movielens.data)) pak::pak("AzKurban-edX-DS/edx.capstone.movielens.data")

library(edx.capstone.movielens.data)
edx <- edx.capstone.movielens.data::edx
final_holdout_test <- edx.capstone.movielens.data::final_holdout_test
```

Now, we have the datasets listed above:
```{r }
summary(edx)
```

```{r }
summary(final_holdout_test)
```

#### `edx` Dataset
\
Let's look into the details of the `edx` dataset:
``` {r echo = TRUE}
str(edx)
```
Note that we have 9000055 rows and six columns in there:  
```{r }
dim_edx <- dim(edx)
print(dim_edx)
```

First, let's note that we have 10677 different movies: 
```{r}
n_movies <- n_distinct(edx$movieId)
print(n_movies)
```
and 69878 different users in the dataset:
```{r}
n_users <- n_distinct(edx$userId)
print(n_users)
```

Finally, we can see that no movies have a rating of 0. Movies are rated from 0.5 to 5.0 in 0.5 increments:
```{r }
#library(dplyr)
s <- edx |> group_by(rating) |>
  summarise(n = n())
print(s)
```

@@

## Methods / Analysis
::: {.noteblock data-latex=""}
All the source code of the R-scripts is available on the project's [GitHub repository](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/tree/main/r/src). 
:::

@@

First, let's define some helper functions for logging and time-measuring features that we will use in our R scripts. Some of them are listed below:

```{r eval=FALSE }
open_logfile <- function(file_name){
  log_file_name <- as.character(Sys.time()) |> 
    str_replace_all(':', '_') |> 
    str_replace(' ', 'T') |>
    str_c(file_name)

@@

::: {.noteblock data-latex=""}
The full source code of these functions is available in the [Logging Helper function section of the Capstone MovieLens Main R Script](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/capstone-movielens.main.R#L1).
:::

### Preparing train and set datasets
\

We will split the `edx` dataset into a training set, which we will use to build and train our models, and a test set in which we will compute the accuracy of our predictions, the way described in [Section 23.1.1 Movielens data](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_ mentioned above[@IDS2_23-1-1].We will also use the _5-Fold Cross Validation_ method the way described in [Section 29.6 Cross validation](https://rafalab.dfci.harvard.edu/dsbook-part-2/ml/resampling-methods.html#cross-validation) of the _Course Textbook_. 
To prepare datasets for processing, we will use the following functions, specifically designed for these operations:

- `make_source_datasets`;
- `init_source_datasets`.

::: {.noteblock data-latex=""}
The full source code of the function listed above is available in the [`Initialize input datasets` section of the `data.helper.functions.R` script on _GitHub_](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/support-functions/data.helper.functions.R#L86)
:::

```{r eval=FALSE, echo=TRUE, include=TRUE}
# Let's ignore the data for users who have not provided at least 100 ratings:
edx100 <- edx |> 
  group_by(userId) |>
  filter(n() >= 100) |>
  ungroup()

### 2025-05-04 reports/capstone-movielens-report.draft4.tex
Draft report updated.

@@

\title{Capstone Movielens Report}
\author{Azamat Kurbanaev}
\date{2025-05-04}

@@

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\subsection{Plot dependency of RMSEs vs Overal Mean Rating
Deviation}\label{plot-dependency-of-rmses-vs-overal-mean-rating-deviation}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{delta =}\NormalTok{ deviation, }
           \AttributeTok{delta.RMSE =}\NormalTok{ deviation.RMSE) }\SpecialCharTok{|\textgreater{}} 
\FunctionTok{tuning.plot}\NormalTok{(}\AttributeTok{title =} \FunctionTok{TeX}\NormalTok{(r}\StringTok{\textquotesingle{}[RMSE as a function of deviation ($\textbackslash{}delta$) from the Overall Mean Rating ($\textbackslash{}hat\{mu\}$)]\textquotesingle{}}\NormalTok{),}
              \AttributeTok{xname =} \StringTok{"delta"}\NormalTok{, }
              \AttributeTok{yname =} \StringTok{"delta.RMSE"}\NormalTok{, }
              \AttributeTok{xlabel =} \FunctionTok{TeX}\NormalTok{(r}\StringTok{\textquotesingle{}[$\textbackslash{}delta$]\textquotesingle{}}\NormalTok{), }
              \AttributeTok{ylabel =} \StringTok{"RMSE"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-1-1.pdf}

@@

\subsubsection{Datasets}\label{datasets}

@@

\FunctionTok{library}\NormalTok{(edx.capstone.movielens.data)}
\NormalTok{edx }\OtherTok{\textless{}{-}}\NormalTok{ edx.capstone.movielens.data}\SpecialCharTok{::}\NormalTok{edx}
\NormalTok{final\_holdout\_test }\OtherTok{\textless{}{-}}\NormalTok{ edx.capstone.movielens.data}\SpecialCharTok{::}\NormalTok{final\_holdout\_test}

\FunctionTok{summary}\NormalTok{(edx)}
\end{Highlighting}
\end{Shaded}

@@

Also, we can see that no movies have a rating of 0. Movies are rated
from 0.5 to 5.0 in 0.5 increments:

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-11-1.pdf}

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-12-1.pdf}

All the source code of the R-scripts is available on the project's
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/tree/main/r/src}{GitHub
repository}.

Let's define some helper functions for logging and time-measuring
features that we will use in our R scripts. Some of them are listed
below:

The full source code of these functions is available in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/capstone-movielens.main.R\#L1}{Logging
Helper function section of the Capstone MovieLens Main R Script}.

\hfill\break
First, let's note that we have 10677 different movies:

@@

Taking into consideration these observations, we came up with a decision
to use the data from the dataset only for users who have provided at
least 100 ratings.

Now let's split the \texttt{edx} dataset into a training set, which we
will use to build and train our models, and a test set in which we will
compute the accuracy of our predictions, the way described in the
\href{https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html\#movielens-data}{Section
23.1.1 Movielens data} of the \emph{Course Textbook} mentioned
above\autocite{IDS2_23-1-1}:

@@

\FunctionTok{print}\NormalTok{(edx100 }\SpecialCharTok{|\textgreater{}} \FunctionTok{summarize}\NormalTok{(}\FunctionTok{n\_distinct}\NormalTok{(userId), }\FunctionTok{n\_distinct}\NormalTok{(movieId)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 1 x 2
##   `n_distinct(userId)` `n_distinct(movieId)`
##                  <int>                 <int>
## 1                24115                 10665
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# For each one of these users, we will split their ratings into 80\% for training }
\CommentTok{\# and 20\% for testing:}

@@

\CommentTok{\# To make sure we don’t include movies in the training set that should not be }
\CommentTok{\# there, we remove entries using the semi\_join function:}
\NormalTok{test\_set }\OtherTok{\textless{}{-}}\NormalTok{ test\_set }\SpecialCharTok{|\textgreater{}} \FunctionTok{semi\_join}\NormalTok{(train\_set, }\AttributeTok{by =} \StringTok{"movieId"}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{as.data.frame}\NormalTok{()}
\FunctionTok{summary}\NormalTok{(test\_set)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      userId         movieId          rating        timestamp            title              genres         
##  Min.   :    8   Min.   :    1   Min.   :0.500   Min.   :8.248e+08   Length:1397040     Length:1397040    
##  1st Qu.:18161   1st Qu.: 1022   1st Qu.:3.000   1st Qu.:9.655e+08   Class :character   Class :character  
##  Median :35666   Median : 2109   Median :3.500   Median :1.058e+09   Mode  :character   Mode  :character  
##  Mean   :35902   Mean   : 4616   Mean   :3.473   Mean   :1.051e+09                                        
##  3rd Qu.:53645   3rd Qu.: 3993   3rd Qu.:4.000   3rd Qu.:1.135e+09                                        
##  Max.   :71565   Max.   :65133   Max.   :5.000   Max.   :1.231e+09
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{train\_set }\OtherTok{\textless{}{-}} \FunctionTok{mutate}\NormalTok{(train\_set, }\AttributeTok{userId =} \FunctionTok{factor}\NormalTok{(userId), }\AttributeTok{movieId =} \FunctionTok{factor}\NormalTok{(movieId))}
\FunctionTok{summary}\NormalTok{(train\_set)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      userId           movieId            rating        timestamp            title              genres         
##  59269  :   5292   356    :  12769   Min.   :0.500   Min.   :8.248e+08   Length:5539951     Length:5539951    
##  67385  :   5088   296    :  12439   1st Qu.:3.000   1st Qu.:9.657e+08   Class :character   Class :character  
##  14463  :   3718   480    :  12390   Median :3.500   Median :1.059e+09   Mode  :character   Mode  :character  
##  68259  :   3228   593    :  12333   Mean   :3.472   Mean   :1.051e+09                                        
##  27468  :   3218   260    :  12142   3rd Qu.:4.000   3rd Qu.:1.135e+09                                        
##  19635  :   3016   1196   :  11586   Max.   :5.000   Max.   :1.231e+09                                        
##  (Other):5516391   (Other):5466292
\end{verbatim}

>>>>

\usepackage{awesomebox}

@@

\title{Capstone Movielens Report}
\author{Azamat Kurbanaev}
\date{2025-05-05}

@@

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\newenvironment{infobox}[1]
  {
  \begin{itemize}
  \renewcommand{\labelitemi}{
    \raisebox{-.7\height}[0pt][0pt]{
      {\setkeys{Gin}{width=3em,keepaspectratio}
        \includegraphics{images/#1}}
    }
  }
  \setlength{\fboxsep}{1em}
  \begin{blackbox}
  \item
  }
  {
  \end{blackbox}
  \end{itemize}
  }

@@

\subsubsection{Datasets Overview}\label{datasets-overview}

@@

\FunctionTok{library}\NormalTok{(edx.capstone.movielens.data)}
\NormalTok{edx }\OtherTok{\textless{}{-}}\NormalTok{ edx.capstone.movielens.data}\SpecialCharTok{::}\NormalTok{edx}
\NormalTok{final\_holdout\_test }\OtherTok{\textless{}{-}}\NormalTok{ edx.capstone.movielens.data}\SpecialCharTok{::}\NormalTok{final\_holdout\_test}
\end{Highlighting}
\end{Shaded}

Now, we have the datasets listed above:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{summary}\NormalTok{(edx)}
\end{Highlighting}
\end{Shaded}

@@

First, let's note that we have 10677 different movies:

@@

Finally, we can see that no movies have a rating of 0. Movies are rated
from 0.5 to 5.0 in 0.5 increments:

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-23-1.pdf}

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-24-1.pdf}

\begin{noteblock}
All the source code of the R-scripts is available on the project's
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/tree/main/r/src}{GitHub
repository}.

\end{noteblock}

First, let's define some helper functions for logging and time-measuring
features that we will use in our R scripts. Some of them are listed
below:

\begin{noteblock}
The full source code of these functions is available in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/capstone-movielens.main.R\#L1}{Logging
Helper function section of the Capstone MovieLens Main R Script}.

\end{noteblock}

We will split the \texttt{edx} dataset into a training set, which we
will use to build and train our models, and a test set in which we will
compute the accuracy of our predictions, the way described in
\href{https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html\#movielens-data}{Section
23.1.1 Movielens data} of the \emph{Course Textbook} mentioned
above\autocite{IDS2_23-1-1}.We will also use the \emph{5-Fold Cross
Validation} method the way described in
\href{https://rafalab.dfci.harvard.edu/dsbook-part-2/ml/resampling-methods.html\#cross-validation}{Section
29.6 Cross validation} of the \emph{Course Textbook}. To prepare
datasets for processing, we will use the following functions,
specifically designed for these operations:

\begin{itemize}
\tightlist
\item
  \texttt{make\_source\_datasets};
\item
  \texttt{init\_source\_datasets}.
\end{itemize}

\begin{noteblock}
The full source code of the function listed above is available in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/support-functions/data.helper.functions.R\#L86}{\texttt{Initialize\ input\ datasets}
section of the \texttt{data.helper.functions.R} script on \emph{GitHub}}

\end{noteblock}

@@

\FunctionTok{print}\NormalTok{(edx100 }\SpecialCharTok{|\textgreater{}} \FunctionTok{summarize}\NormalTok{(}\FunctionTok{n\_distinct}\NormalTok{(userId), }\FunctionTok{n\_distinct}\NormalTok{(movieId)))}

\CommentTok{\# For each one of these users, we will split their ratings into 80\% for training }
\CommentTok{\# and 20\% for testing:}

@@

\CommentTok{\# To make sure we don’t include movies in the training set that should not be }
\CommentTok{\# there, we remove entries using the semi\_join function:}
\NormalTok{test\_set }\OtherTok{\textless{}{-}}\NormalTok{ test\_set }\SpecialCharTok{|\textgreater{}} \FunctionTok{semi\_join}\NormalTok{(train\_set, }\AttributeTok{by =} \StringTok{"movieId"}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{as.data.frame}\NormalTok{()}
\FunctionTok{summary}\NormalTok{(test\_set)}

\NormalTok{train\_set }\OtherTok{\textless{}{-}} \FunctionTok{mutate}\NormalTok{(train\_set, }\AttributeTok{userId =} \FunctionTok{factor}\NormalTok{(userId), }\AttributeTok{movieId =} \FunctionTok{factor}\NormalTok{(movieId))}
\FunctionTok{summary}\NormalTok{(train\_set)}
\end{Highlighting}
\end{Shaded}
